Embedding intuitionistic into classical logic

### 2022-10-05 Notes/Embedding intuitionistic into classical logic/EIICL.tex
Defensio

\usepackage{setspace}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{ wasysym }
\usepackage{ stmaryrd }
\usepackage{ mathpartir }
\usepackage{bussproofs}
\usepackage{enumerate}
\usepackage{tikz-qtree}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{arrows,shapes,quotes}
\usepackage{stackengine}
\usepackage{footmisc}
\usepackage{wrapfig}
\usepackage[a4paper,top=2.5cm,bottom=3.5cm,left=2.5cm,right=2.5cm,headsep=10pt]{geometry}
\usepackage[toc,page]{appendix}
\usepackage{listings}
\lstset{
	language=bash,
	basicstyle=\ttfamily
}

@@

\theoremstyle{definition}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{definition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{definition}
\newtheorem{remark}[theorem]{Remark}

\begin{document}

	\pagenumbering{gobble}
	\begin{flushright}
		% https://public.univie.ac.at/en/downloads/logos/
		\includegraphics[width=7cm]{RZ_Logo_Uni_sw}
	\end{flushright}	
	\vskip20mm
	\begin{center}
		\huge\textbf{MASTERARBEIT / MASTER'S THESIS}
		\vskip15mm
		\normalsize{Titel der Masterarbeit / Title of the Master's Thesis}

@@

		\Large\textbf{Alexander Pluska, B.Sc.} % Insert your name in this line
		\vskip15mm
		\normalsize{angestrebter akademischer Grad / in partial fulfillment of the requirements for the degree of}

		\Large\textbf{Master of Science (M.Sc.)}

		\small
	\end{center}
	\vskip15mm
	\begingroup
	\renewcommand{\arraystretch}{1.2}
	\begin{flushleft}
		Wien, 2022 / Vienna 2022 % Insert graduation year in this line
		\vfill
		\addtolength{\tabcolsep}{-6pt} 
		\begin{table}[h!]
			\begingroup
			\renewcommand*{\arraystretch}{1.2}
			%\renewcommand*{\linespacing}{1}
			\renewcommand\baselinestretch{1}
			\small
			\begin{tabular}{p{7cm} p{10cm}}
				Studienkennzahl lt. Studienblatt / \newline degree programme code as it appears on \newline the student record sheet: &A 066 821 \\
				Studienrichtung lt. Studienblatt / \newline degree programme as it appears on \newline the student record sheet:& Mathematics \\
				Betreut von / Supervisor: & Univ.-Prof. Matthias Aschenbrenner \\ % Insert your program/supervisor here
				Mitbetreut von / Co-Supervisor: & Ass.-Prof. Florian Zuleger \\ 
			\end{tabular}
			\endgroup
		\end{table}
		\addtolength{\tabcolsep}{6pt} 
	\end{flushleft}	
	\endgroup

@@

		\
		\vfill
		\onehalfspacing
		\begin{center}
		\bfseries Abstract		
		\end{center}
		The famous double negation translation~\cite{glivenko1929quelques, godel1933intuitionistischen} establishes an embedding from classical into intuitionistic logic. Curiously the reverse direction has not been covered in literature. We present an effective embedding from intuitionistic into classical logic, both in the propositional and first-order case, as well as an effective embedding of intuitionistic propositional logic into quantified boolean formulas.

		Furthermore we implement a practical system that takes intuitionistic first-order problems as an input and performs our transformation. This allows the use of classical theorem provers for establishing intuitionistic validity. We benchmark our implementation using the vampire theorem prover~\cite{kovacs2013first} on the ILTP problem set~\cite{iltp}.
		%The key notion used in the proofs is the existence of counter-models - an analogon to the satisfiability of the negation in classical logic, which is not equivalent to invalidity in the intuitionistic case.  We hope that this will allow leveraging the tremendous recent progress in automated reasoning in classical logic for checking intuitionistic validity, for which there has not been much progress in recent decades.

		Finally we discuss how the generated classical proofs of the transformed problems can be translated to intuitionistic proofs of the original problem. All in all this establishes a novel approach to theorem proving for intuitionistic logic.

@@

	I want to thank my Matthias for accommodating my particular interests by agreeing to supervise a thesis which lies outside of his main area of research and bearing with the organizational difficulties. I want to thank Florian who graciously agreed to co-supervise a thesis at a different university. He has provided me with a fascinating topic as well as plenty of his time, expertise and thorough feedback. Finally I want to thank my family, in particular my parents Stefania and Cezary, who have instilled in me the joy of learning, provided all the love and help I could ask for and invariably supported me along my journey.

@@

	\newpage
	\
	\singlespacing
	\newgeometry{top=1cm}
	\tableofcontents
	\restoregeometry

@@

	The interest in intuitionistic logic has lead to the development of a number of automated theorem proving systems for propositional as well as for predicate logic and a collection of benchmark problems (see e.g. the
	ILTP library website~\cite{iltp}).
	However, progress in automated theorem proving for intuitionistic logic has been slow, whereas reasoners for classical logic have made tremendous progress, see e.g. the TPTP~\cite{casc} and SAT~\cite{satc} competitions.
	%, e.g., the first-order provers~\cite{kovacs2013first, schulz2002brainiac, korovin2008iprover}
	%Comparing benchmarks for classical and intuitionistic automated theorem proving, e.g. TPTP~\cite{casc} and , we see that classical provers currently have a far better success rate. Furthermore all the most popular first-order utilize classical logic.
	This difference can be partially explained by fundamental differences between the logics.
	First of all, determining intuitionistic validity is computationally harder, i.e. in the propositional case intuitionistic validity is \verb+PSPACE+-complete~\cite{statman1979intuitionistic} whereas classical validity is \verb+coNP+-complete~\cite{cook1971complexity}.
	A further advantage of classical logic is the existence of calculi that are particularly suitable for automation such as superposition~\cite{bachmair2001resolution}, which rely on the existence of convenient normal forms such as CNF, and the duality between validity and satisfiability, i.e. to show the validity of a formula it suffices to show the unsatisfiability of the negated formula.
	While some (albeit more complex) normal forms also exist  for intuitionistic logic, crucially the duality between validity and satisfiability of the negation does not hold.
	Therefore most dedicated intuitionistic theorem provers~\cite{mclaughlin2009efficient, tammet1996resolution} use the naive inverse method, i.e. directly search for a cut-free proof by applying the rules from some proof calculus inversely, enhanced by search strategies such as focussing and polarization. This approach generally leads to a much more complex search and is therefore difficult to implement efficiently.
	Finally, we add that in contrast to intuitionistic provers a tremendous amount of work has been put into optimizing provers for classical logic, in particular for the propositional case, i.e. SAT-solvers.

	With this work we want to propose a novel approach for intuitionistic theorem proving that leverages the progress in classical theorem proving, that is for each formula $\varphi$:
	\begin{itemize}
		\item Give a formula $\varphi^\#$ that is classically valid if and only if $\varphi$ is intuitionistically valid.
		\item Use a state-of-the-art classical prover to establish the validity/invalidity of $\varphi^\#$.
		\item Transform the output proof/counter-model of $\varphi^\#$ to one of $\varphi$.
	\end{itemize}
	The most challenging part of this approach is giving the translation of $\varphi$ to $\varphi^\#$.
	Interestingly, the reverse direction, the famous double-negation translation, has long been established and goes back to Glivenko~\cite{glivenko1929quelques} in the propositional case, and to G\"odel~\cite{godel1933intuitionistischen} and Gentzen~\cite{gentzen1936widerspruchsfreiheit} in the first-order case. In the propositional case it is particularly simple: $\varphi$ is classically valid if and only if $\neg\neg\varphi$ is intuitionistically valid. Intuitively the translation collapses for each subformula $\psi$ of $\varphi$ the truth values of $\psi$ and $\neg\neg\psi$, which are classically but not intuitionistically equivalent. This gives us a first idea why the reverse direction is perhaps more difficult: We need to expand the truth values of $\psi$ and $\neg\neg\psi$, i.e. if they both occur in $\varphi$, we must have a way to (classically) assign different truth values to their respective counterparts in $\varphi^\#$. In particular, this necessitates the introduction of new propositional variables in our procedure, which already marks a big difference to the double negation translation.
	%
	While establishing the translation of formulas we also perform an effective translation of counter-models, i.e. for each intuitionistic counter-model of $\varphi$ we effectively construct a classical counter-model of $\varphi^\#$ and vice versa.
	We note that the existence of counter-models is a key notion that forms a proper dual to validity --- whereas the satisfiability of the negation is not necessary for invalidity (in contrast to classical logic, where it is a proper dual to validity).
	Transforming and reducing counter-models to a normal form is also what ultimately enables our translation. Apart from a translation of counter-models we also explicitly describe how to equivalently transform classical proofs of $\varphi^\#$ to intuitionistic proofs of $\varphi$, in particular resolution proofs as given by theorem provers like vampire~\cite{Kov_cs_2013}.

	As a final contribution we implement our translation in the Rust programming language. It takes a first-order problem in the tptp format~\cite{tptp} and outputs the translated problem in the same format. The code is published on github~\cite{implementation}. We then benchmark our implementation using the ILTP problem set~\cite{iltp}, i.e. we translate all problems in the set and then run the vampire theorem prover~\cite{Kov_cs_2013} on the translated problems. Our approach performs on par with most existing approaches for intuitionistic theorem provers, but comes short of the state-of-the art leaders. As there is still a lot of room for optimization this is a hopeful sign.

@@

	The main content is organized in chapters 3-6. Chapter 3 collects prerequisite definitions and serves to fix our notation and recall some important facts. Since this overview already contains some technical results it may be consulted for reference. Chapter 4 is the meat of this thesis. In particular it features a complete description of our embedding of intuitionistic into classical logic as well as proving its correctness. In chapter 5 we present our implementation and benchmark it. Finally in chapter 6 we explicitly describe how to translate the generated proofs and counter-models which in principle directly follows from the results of section 4. 	We shall now for each chapter give an overview of our main results and highlight the key arguments.

	\section{Embedding Intuitionistic into Classical Logic}

@@

	Before our main transformation we employ a preprocessing step akin to the Tseytin transformation~\cite{tseitin1983complexity}, which is a popular pre-processing step in classical automated reasoning:
	It gives an equisatisfiable sentence (over an extended set of propositions) in conjunctive normal form.
	Eliminating all implications, however, is not possible in intuitionistic logic since $A\to B$ is not equivalent to $\neg A\vee B$.
	Still we propose a similar transformation.
	A notable feature of our transformation is that all non-classical content is encapsulated in formulas of type $(A\to B)\to C$, i.e. if there are no such formulas in the transformed formula then the classical validity of the transformed formula immediately implies intuitionistic validity.

	\begin{theorem}\label{thm:Tseytin}
		For every propositional formula $\varphi$ there effectively are an atom $P$ and a set of $\mathcal S$ of formulas (over an extended set of propositions) of one of the forms\\
		$\text{ \hspace{1cm}} A\to (B\wedge C), (A\wedge B)\to C, A\to (B\vee C), (A\vee B)\to C, (A\to B)\to C,$\\
		for atomic $A, B, C$, such that $\varphi$ is intuitionistically valid if and only if $\bigwedge\mathcal S\to P$ is intuitionistically valid. The time complexity of the transformation is linear in the input size.
	\end{theorem}

	The main transformation then proceeds in three steps:
	%
	1) We encode as a first-order sentence that the considered formula holds for every Kripke frame.
	Since Kripke frames for propositional logic over a fixed alphabet of propositional variables form a first-order theory this step is rather straightforward.
	%
	2) Next we eliminate some quantifiers via Herbrandization. The only quantifiers eliminated occur in formulas derived from formulas of type $(A\to B)\to C$.
	%
	3) We can then argue that if there is a counter-model for the resulting formula then there is a counter-model with a certain Kripke frame of bounded size completely determined by the number of formulas of type $(A\to B)\to C$.
	This allows us to eliminate the remaining quantifiers by simply enumerating the worlds in that Kripke frame.

@@

	\begin{theorem}
		\label{thm:reduction-propositional}
		Let $\mathcal S$ be as in Theorem~\ref{thm:Tseytin} and $\mathcal F_\to\subseteq\mathcal S$ denote the subset of formulas of the form $(A\to B)\to C$ and $\Lambda$ denote the set of sequences without repetition over $\mathcal F_\to$. For each atom $A$ and $k\in\Lambda$ consider a new atom $A^k$. Obtain $\mathcal S^\#$ by including the following formulas:
		\begin{itemize}
			\item $A^k\to A^{k\psi}$ for each atom $A$ occurring in $\mathcal S$, $k\in\Lambda$ and $\psi\in\mathcal F_\to$ not occurring in $k$.
			\item $A^k\to (B^k\circ C^k)$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
			\item $(A^k\circ B^k)\to C^k$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
			\item $(A^{k\psi}\to B^{k\psi})\to C^k$ for $\psi = (A\to B)\to C\in\mathcal S$, $k\in\Lambda$ if $\psi$ does not occurr in $k$.
		\end{itemize}
		Then, $\bigwedge S\to P$ is intuitionistically valid iff $\bigwedge S^\#\to P^\epsilon$ is classically valid, where $\epsilon$ denotes the empty sequence. The size of $\mathcal S^\#$ is in $\mathcal O(|\mathcal S|\cdot2^{|\mathcal F_\to|\cdot\log(|\mathcal F_\to|)})$. Futhermore, there is an effective procedure for translating counter-models between the sentences.
	\end{theorem}

	Instead of explicitly enumerating the worlds of a counter-example of bounded size (as in the reduction stated in the above theorem), we can also do this enumeration implicitly with a quantified boolean formula (QBF).
	We can show that this QBF is satisfiable if and only if there exists a counter-model for our original formula:

@@

	We then move to the first-order case. Here the construction is a bit more involved but the underlying approach is the same.
	Again we first perform a Tseytin-like transformation.
	Here the non-classical content is encapsulated by formulas of type $(A\to B)\to C$ and $\forall xA\to B$.
	The details of this translation are given in Section~\ref{section:tseytin}.
	We then proceed similarly to the propositional case.
	Note however that one difficulty arises from the fact that our (classical) domain will now contain on the one hand worlds in the Kripke frame but on the other hand also proper domain elements.
	We resolve this apparent conflict by introducing a special binary predicate $E$, inspired by~\cite{iemhoff2010eskolemization}, encoding which domain elements exists at which world.
	The main transformation then again proceeds in three steps:
	Step 1) and 2) are analogous to the propositional case.
	The real difference arises in step 3):
	While we can also establish the existence of a canonic counter-model whose frame only depends on formulas of type $(A\to B)\to C$ and $\forall xA\to B$, the size of this counter-model is countably infinite in general.
	This is not surprising since certain intuitionistically invalid formulas like the double negation shift $\forall x\neg\neg A(x)\to \neg\neg\forall x A(x)$ don't have finite counter-models.
	Therefore we are not able to eliminate the $\forall$-quantifiers associated with the Kripke semantics.
	However since our translation targets first-order logic, we believe that the introduction of these quantifiers is acceptable.
	%
	Summarizing, we obtain the following result:

@@

	In this section we describe our system that implements the translation from the previous section. The rust source code can be found online~\cite{implementation}. We then elaborate our benchmarking process which utilizes the ILTP problem set~\cite{iltp} and the vampire theorem prover~\cite{Kov_cs_2013}.

	\begin{figure}[h]\label{results}
		$$\begin{matrix}
			&\text{AGT}&\text{ALG}&\text{COM}&\text{CSR}&\text{GEO}&\text{GRA}&\text{GRP}&\text{HAL}\\
			\text{Proven}&5&170&0&0&58&3&4&2\\
			\text{Disproven}&0&0&0&0&0&0&0&\\
			\text{Timeout}&47&29&3&29&107&74&1&16
		\end{matrix}$$
		$$\begin{matrix}
			&\text{KRS}&\text{LCL}&\text{MGT}&\text{MSC}&\text{NLP}&\text{NUM}&\text{PLA}&\text{PUZ}\\
			\text{Proven}&0&0&30&1&9&1&9&2\\
			\text{Disproven}&0&0&0&0&12&0&0&0\\
			\text{Timeout}&5&9&128&3&69&1&237&82
		\end{matrix}$$
		$$\begin{matrix}
			&\text{SET}&\text{SWC}&\text{SWV}&\text{SYN}&\text{TOP}&\text{GEJ}&\text{GPJ}&\text{SYJ}\\
			\text{Proven}&0&5&27&1&64&65&54&1\\
			\text{Disproven}&6&1&297&423&154&108&298&2
		\end{matrix}$$
		\caption{Performance on problem sections of the ILTP set.}
	\end{figure}

	Our implementation performs similar to existing dedicated provers for intuitionistic logic, but falls short of the state-of-the-art. Since there is still a lot of possible optimizations left for our translation this is a hopeful first sign.

	\section{Translation of proofs and counter-models}

	In the final section we give an explicit mechanism how to transform the proofs and counter-models for translated formula $\varphi^\#$ into intuitionistic proofs/counter-models of the original formula $\varphi$. The existence of these transformations is clear from the methodology of chapter~\ref{chapter:eiicl}, however they are hidden in the proofs and here we make them explicit. The two main results are as follows:
	\begin{theorem}
		There exists an effective linear time procedure that converts any classical counter-model for a translated formula $\varphi^\#$ into a intuitionistic counter-model for the original formula $\varphi$.
	\end{theorem}
	\begin{theorem}
		There exists an effective linear time procedure that converts any classical proof of a translated formula $\varphi^\#$ into a intuitionistic proof for the original formula $\varphi$.
	\end{theorem}

	Given the arguments of Chapter~\ref{chapter:eiicl} the translation of counter-models is more or less just a compilation of proofs. On the other hand the explicit translation of proofs does require some work, since to that point all arguments were semantic and this requires a syntactic viewpoint.

>>>>

\usepackage{setspace}
\usepackage{caption}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{ wasysym }
\usepackage{ stmaryrd }
\usepackage{ mathpartir }
\usepackage{bussproofs}
\usepackage{enumerate}
\usepackage{tikz-qtree}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{arrows,shapes,quotes}
\usepackage{stackengine}
\usepackage{footmisc}
\usepackage{wrapfig}
\usepackage[a4paper,top=3cm,bottom=3.8cm,left=3cm,right=3cm,headsep=20pt]{geometry}
\usepackage[toc,page]{appendix}
\usepackage{listings}
\lstset{
	language=bash,
	basicstyle=\ttfamily
}

@@

\theoremstyle{definition}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{definition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{definition}
\newtheorem{remark}[theorem]{Remark}
\begin{document}

	\pagenumbering{gobble}
	\newgeometry{top=2cm, right=2.5cm, left=2.5cm, bottom=2.5cm}
	\begin{flushright}
		% https://public.univie.ac.at/en/downloads/logos/
		\includegraphics[width=7cm]{RZ_Logo_Uni_sw}
	\end{flushright}	
	\vskip20mm
	\begin{center}
		\huge\textbf{MASTERARBEIT / MASTER'S THESIS}
		\vskip15mm
		\normalsize{Titel der Masterarbeit / Title of the Master's Thesis}

@@

		\Large\textbf{Alexander Pluska, BSc} % Insert your name in this line
		\vskip15mm
		\normalsize{angestrebter akademischer Grad / in partial fulfillment of the requirements for the degree of}

		\Large\textbf{Master of Science (MSc)}

		\small
	\end{center}
	\vskip15mm
	\begingroup
	\renewcommand{\arraystretch}{1.2}
	\begin{flushleft}
		Wien, 2022 / Vienna, 2022 % Insert graduation year in this line
		\vfill
		\addtolength{\tabcolsep}{-6pt} 
		\begin{table}[h!]
			\begingroup
			\renewcommand*{\arraystretch}{1.2}
			%\renewcommand*{\linespacing}{1}
			\renewcommand\baselinestretch{1}
			\small
			\begin{tabular}{p{7cm} p{10cm}}
				Studienkennzahl lt. Studienblatt / \newline degree programme code as it appears on \newline the student record sheet: &UA 066 821 \\
				Studienrichtung lt. Studienblatt / \newline degree programme as it appears on \newline the student record sheet:& Masterstudium Mathematik\\
				Betreut von / Supervisor: & Univ.-Prof. Matthias Aschenbrenner, PhD
			\end{tabular}
			\endgroup
		\end{table}
		\addtolength{\tabcolsep}{6pt} 
	\end{flushleft}	
	\endgroup	
	\restoregeometry

@@

		\
		\vfill
		\onehalfspacing
		\begin{center}
		\bfseries Abstract		
		\end{center}
		The famous double negation translation~\cite{glivenko1929quelques, godel1933intuitionistischen} establishes an embedding from classical into intuitionistic logic. Curiously, the reverse direction has not been covered in literature. We present an effective embedding from intuitionistic into classical logic, both in the propositional and first-order case, as well as an effective embedding of intuitionistic propositional logic into quantified boolean formulas.

		Furthermore, we implement a system that takes intuitionistic first-order problems in the tptp file format as an input and performs our transformation. This allows the use of classical theorem provers for checking intuitionistic validity. We benchmark our implementation using the Vampire theorem prover~\cite{kovacs2013first} on the ILTP problem set~\cite{iltp}.

		Finally, we discuss how the generated classical proofs of the transformed problem can be translated back into intuitionistic proofs of the original problem. All in all, this establishes a novel approach to theorem proving for intuitionistic logic and provides a first proof of concept.

@@

	I want to thank Matthias for accommodating my particular interests by agreeing to supervise a thesis which lies outside his main area of research and bearing with the organizational difficulties.

	I sincerely appreciate Florian's guidance, who graciously agreed to co-supervise a thesis at a different university. He has provided me with a fascinating topic as well as plenty of his time, expertise and thorough feedback.

	Finally, I am incredibly grateful to my family, in particular my parents Stefania and Cezary, who have instilled in me the joy of learning, provided all the love and help I could ask for and invariably supported me along my journey.

@@

	\newpage
	\
	\singlespacing
	\newgeometry{top=1cm, bottom=2cm}
	\tableofcontents
	\restoregeometry

@@

	The interest in intuitionistic logic has led to the development of a number of automated theorem proving systems for propositional as well as for predicate logic and a collection of benchmark problems (see e.g. the
	ILTP library website~\cite{iltp}).
	However, progress in automated theorem proving for intuitionistic logic has been slow, whereas reasoners for classical logic have made tremendous progress, see e.g. the TPTP~\cite{casc} and SAT~\cite{satc} competitions.
	This difference can be partially explained by fundamental differences between the logics.
	Foremost, determining intuitionistic validity is computationally harder, i.e. in the propositional case intuitionistic validity is \verb+PSPACE+-complete~\cite{statman1979intuitionistic} whereas classical validity is \verb+coNP+-complete~\cite{cook1971complexity}.
	A further advantage of classical logic is the existence of calculi that are particularly suitable for automation such as superposition~\cite{bachmair2001resolution}, which rely on the existence of convenient normal forms such as CNF, and the duality between validity and satisfiability, i.e. to show the validity of a formula it suffices to show the unsatisfiability of the negated formula.
	While some (albeit more complex) normal forms also exist  for intuitionistic logic, crucially the duality between validity and satisfiability of the negation does not hold.
	Therefore, most dedicated intuitionistic theorem provers~\cite{mclaughlin2009efficient, tammet1996resolution} use the naive inverse method, i.e. direct search for a cut-free proof by applying the rules from some proof calculus inversely, enhanced by search strategies such as focussing and polarization. This approach generally leads to a much more complex search and is therefore difficult to implement efficiently.
	Finally, we add that in contrast to intuitionistic provers a tremendous amount of work has been put into optimizing provers for classical logic, in particular for the propositional case, i.e. SAT-solvers.

	With this work we want to propose a novel approach for intuitionistic theorem proving that leverages the progress in classical theorem proving, that is for each formula $\varphi$:
	\begin{itemize}
		\item Give a formula $\varphi^\#$ that is classically valid if and only if $\varphi$ is intuitionistically valid.
		\item Use a state-of-the-art classical prover to establish the validity/invalidity of $\varphi^\#$.
		\item Transform the generated proof/counter-model of $\varphi^\#$ to one of $\varphi$.
	\end{itemize}
	The most challenging part of this approach is giving the translation of $\varphi$ to $\varphi^\#$.
	Interestingly, the reverse direction, the famous double-negation translation, has long been established and goes back to Glivenko~\cite{glivenko1929quelques} in the propositional case, and to G\"odel~\cite{godel1933intuitionistischen} and Gentzen~\cite{gentzen1936widerspruchsfreiheit} in the first-order case. In the propositional case, it is particularly simple: $\varphi$ is classically valid if and only if $\neg\neg\varphi$ is intuitionistically valid. Intuitively, the translation collapses for each subformula $\psi$ of $\varphi$ the truth values of $\psi$ and $\neg\neg\psi$, which are classically but not intuitionistically equivalent. This gives us a first idea why the reverse direction is perhaps more difficult: We need to expand the truth values of $\psi$ and $\neg\neg\psi$, i.e. if they both occur in $\varphi$, we must have a way to (classically) assign different truth values to their respective counterparts in $\varphi^\#$. In particular, this necessitates the introduction of new propositional variables in our procedure, which already marks a big difference to the double negation translation.

	While establishing the translation of formulas we also perform an effective translation of counter-models, i.e. for each intuitionistic counter-model of $\varphi$ we effectively construct a classical counter-model of $\varphi^\#$ and vice versa.
	We note that the existence of counter-models is a key notion that forms a proper dual to validity --- whereas the satisfiability of the negation is not necessary for invalidity (in contrast to classical logic, where it is a proper dual to validity).
	Transforming and reducing counter-models to a normal form is also what ultimately enables our translation. Apart from a translation of counter-models in the final chapter, we also explicitly describe how to equivalently transform classical proofs of $\varphi^\#$ to intuitionistic proofs of $\varphi$.

	As a final contribution, we implement our translation in the Rust programming language. Our implementation transforms a first-order problem in the tptp format~\cite{tptp} and outputs the translated problem in the same format. The code is published on GitHub~\cite{implementation}. We then benchmark our implementation using the ILTP problem set~\cite{iltp}, i.e. we translate all problems in the set and then run the Vampire theorem prover~\cite{Kov_cs_2013} on the translated problems. Our approach performs on par with existing approaches for intuitionistic theorem proving, but comes short of the state-of-the art. As there is still a lot of room for optimization, this is a hopeful first sign.

@@

	The main content is organized in chapters 3 to 6. Chapter 3 collects prerequisite definitions and serves to fix our notation and recall some important facts. Chapter 4 is the meat of this thesis. In particular, it features a complete description of our embedding of intuitionistic into classical logic as well as proving its correctness. In chapter 5 we present the implementation of the embedding and the results of our benchmark. Finally, in chapter 6 we explicitly describe how to translate the generated proofs and counter-models which in principle directly follows from the results of section 4. 	We shall now for each chapter give an overview of our main results and highlight the key arguments.

	\section{Embedding intuitionistic into classical logic}

@@

	Before our main transformation we employ a preprocessing step akin to the Tseytin transformation~\cite{tseitin1983complexity}, which is a popular pre-processing step in classical automated reasoning:
	It gives an equisatisfiable sentence (over an extended set of propositions) in conjunctive normal form.
	Eliminating all implications, however, is not possible in intuitionistic logic since $A\to B$ is not equivalent to $\neg A\vee B$.
	Still, we propose a similar transformation.
	A notable feature of our transformation is that all non-classical content is encapsulated in formulas of type $(A\to B)\to C$, i.e. if there are no such formulas in the transformed formula then the classical validity of the transformed formula immediately implies intuitionistic validity and no further processing is needed.

	\begin{theorem}\label{thm:Tseytin1}
		For every propositional formula $\varphi$ there effectively are an atom $P$ and a set of $\mathcal S$ of formulas (over an extended set of propositions) of one of the forms\\
		$\text{ \hspace{1cm}} A\to (B\wedge C), (A\wedge B)\to C, A\to (B\vee C), (A\vee B)\to C, (A\to B)\to C,$\\
		for atomic $A, B, C$, such that $\varphi$ is intuitionistically valid if and only if $\bigwedge\mathcal S\to P$ is intuitionistically valid. The time complexity of the transformation is linear in the input size.
	\end{theorem}

	The main transformation then proceeds in three steps:
	%
	1) We encode as a first-order sentence that the considered formula holds for every Kripke frame.
	Since Kripke frames for propositional logic over a fixed alphabet of propositional variables form a first-order theory, this step is rather straightforward.
	%
	2) Next we eliminate some quantifiers via Herbrandization. The only quantifiers eliminated occur in formulas derived from formulas of type $(A\to B)\to C$.
	%
	3) We can then argue that if there is a counter-model for the resulting formula, then there is a counter-model with a certain Kripke frame of bounded size completely determined by the number of formulas of type $(A\to B)\to C$.
	This allows us to eliminate the remaining quantifiers by simply enumerating the worlds in that Kripke frame.

@@

	\begin{theorem}
		\label{thm:reduction-propositional}
		Let $\mathcal S$ be as in Theorem~\ref{thm:Tseytin1} and $\mathcal F_\to\subseteq\mathcal S$ denote the subset of formulas of the form $(A\to B)\to C$ and $\Lambda$ denote the set of sequences without repetition over $\mathcal F_\to$. For each atom $A$ and $k\in\Lambda$ consider a new atom $A^k$. Obtain $\mathcal S^\#$ by including the following formulas:
		\begin{itemize}
			\item $A^k\to A^{k\psi}$ for each atom $A$ occurring in $\mathcal S$, $k\in\Lambda$ and $\psi\in\mathcal F_\to$ not occurring in $k$.
			\item $A^k\to (B^k\circ C^k)$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
			\item $(A^k\circ B^k)\to C^k$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
			\item $(A^{k\psi}\to B^{k\psi})\to C^k$ for $\psi = (A\to B)\to C\in\mathcal S$, $k\in\Lambda$ if $\psi$ does not occur in $k$.
		\end{itemize}
		Then, $\bigwedge S\to P$ is intuitionistically valid iff $\bigwedge S^\#\to P^\epsilon$ is classically valid, where $\epsilon$ denotes the empty sequence. The size of $\mathcal S^\#$ is in $\mathcal O(|\mathcal S|\cdot2^{|\mathcal F_\to|\cdot\log(|\mathcal F_\to|)})$. Furthermore, there is an effective procedure for translating counter-models between the sentences.
	\end{theorem}

	Instead of explicitly enumerating the worlds of a counter-example of bounded size (as in the reduction stated in the above theorem), we can also do this enumeration implicitly with a quantified boolean formula (QBF).
	We can show that this QBF is satisfiable iff there exists a counter-model for our original formula:

@@

	We then move to the first-order case. Here the construction is a bit more involved, but the underlying approach is the same.
	Again, we first perform a Tseytin-like transformation.
	Here the non-classical content is encapsulated by formulas of type $(A\to B)\to C$ and $\forall xA\to B$.
	The details of this translation are given in Section~\ref{section:tseytin}.
	We then proceed similarly to the propositional case.
	Note however that one difficulty arises from the fact that our (classical) domain will now contain on the one hand worlds in the Kripke frame, but on the other hand also proper domain elements.
	We resolve this apparent conflict by introducing a special binary predicate $E$, inspired by~\cite{iemhoff2010eskolemization}, encoding which domain elements exists at which world.
	The main transformation then again proceeds in three steps:
	Step 1) and 2) are analogous to the propositional case.
	The real difference arises in step 3):
	While we can also establish the existence of a canonic counter-model whose frame only depends on formulas of type $(A\to B)\to C$ and $\forall xA\to B$, the size of this counter-model is countably infinite in general.
	This is not surprising, since certain intuitionistically invalid formulas like the double negation shift $\forall x\neg\neg A(x)\to \neg\neg\forall x A(x)$ don't have finite counter-models.
	Therefore, we are not able to eliminate the $\forall$-quantifiers associated with the Kripke semantics.
	However, since our translation targets first-order logic, we believe that the introduction of these quantifiers is acceptable.

@@

	In this section, we describe our system that implements the translation from the previous section. The rust source code can be found online~\cite{implementation}. We then elaborate our benchmarking process, which utilizes the ILTP problem set~\cite{iltp} and the Vampire theorem prover~\cite{Kov_cs_2013}.

	\begin{figure}[h]\label{results}
		$$\begin{matrix}
			&\text{AGT}&\text{ALG}&\text{COM}&\text{CSR}&\text{GEJ}&\text{GEO}&\text{GPJ}&\text{GRA}\\
			\text{Proven}&5&170&0&0&58&3&4&2\\
			\text{Disproven}&0&0&0&0&0&0&0&0\\
			\text{Timeout}&47&29&3&29&107&74&1&16
		\end{matrix}$$
		$$\begin{matrix}
			&\text{GRP}&\text{HAL}&\text{KRS}&\text{LCL}&\text{MGT}&\text{MSC}&\text{NLP}&\text{NUM}\\
			\text{Proven}&0&0&30&1&9&1&9&2\\
			\text{Disproven}&0&0&0&0&12&0&0&0\\
			\text{Timeout}&5&9&128&3&69&1&237&82
		\end{matrix}$$
		$$\begin{matrix}
			&\text{PLA}&\text{PUZ}&\text{SET}&\text{SWC}&\text{SWV}&\text{SYJ}&\text{SYN}&\text{TOP}\\
			\text{Proven}&0&5&27&1&64&65&54&1\\
			\text{Disproven}&0&1&0&0&2&3&13&0\\
			\text{Timeout}&6&1&297&422&154&108&298&2
		\end{matrix}$$
		\caption{Performance on problem sections of the ILTP set.}
	\end{figure}

	Our implementation performs similar to existing dedicated provers for intuitionistic logic, but falls short of the state-of-the-art. Since there is still a lot of possible optimizations left for our translation, this is a hopeful first sign.

	\section{Translation of counter-models and proofs}

	In the final section, we give an explicit mechanism how to transform the proofs and counter-models for translated formula $\varphi^\#$ into intuitionistic proofs/counter-models of the original formula $\varphi$. The existence of these transformations is clear from the methodology of chapter~\ref{chapter:eiicl}, however, they are hidden in the proofs and here we make them explicit. The two main results are as follows:
	\begin{theorem}
		There exists an effective procedure that converts any classical counter-model for a translated formula $\varphi^\#$ into an intuitionistic counter-model for the original formula $\varphi$.
	\end{theorem}
	\begin{theorem}
		There exists an effective procedure that converts any classical proof of a translated formula $\varphi^\#$ into a intuitionistic proof of the original formula $\varphi$.
	\end{theorem}

	Given the arguments of Chapter~\ref{chapter:eiicl} the translation of counter-models is more or less just a compilation of proofs. On the other hand, the explicit translation of proofs does require some work, since to that point all arguments were semantic and this requires a syntactic viewpoint.

	\chapter{Preliminaries}
	First, let us recapitulate the most important definitions and some facts about them. This will serve to fix our notation as well.
